NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2018-04-20 Manuscript/manuscriptPS.tex
Update manuscript

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/comparisonTablePS.pdf}
  \caption{\label{comparisonTablePS}
  Rough subjective ranking of force fields based on Figure~\ref{HGorderParametersPS}. Here ÒMÓ indicates a magnitude problem, ÒFÓ a forking problem; letter size increases with problem severity. Color scheme: Òwithin experimental errorÓ (dark green), Òalmost within experimental errorÓ (light green), Òclear deviation from experimentsÓ (light red), and Òmajor deviation from experimentsÓ (dark red). The $\Sigma$-column shows the total deviation of the force field, when individual carbons are given weights of 0 (matches experiment), 1, 2, and 4 (major deviation). For full details of the assessment, see Supplementary Information.
  }
  \todo{Issue about possible updates to this plot: https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/4}
\end{figure}

The headgroup order parameters of DOPS and POPS bilayers 
from different simulation models are compared with the experimental
data in Fig. \ref{HGorderParametersPS}. In line with the previous
study for PC lipids \cite{botan15}, the glycerol backbone order 
parameters in CHARMM36 roughly agree with the experimental data,
while significant discrepancies for glycerol backbone carbons
are observed in other models.
\todo{Discussion to be finished once all the results are in the plot.}

@@

In order to compare the sodium interaction with PS containing membranes between
experiments and simulations, Fig. \ref{PSresponseTONaCl} shows the order parameters of the
$\alpha$ carbon of PS headgroup measured from the mixture with PC as a
function of added NaCl \cite{roux86} together with the resuls from CHARMM36 model. 
\begin{figure*}[]
  \centering
  \includegraphics[width=18.0cm]{../Figs/CHANGESwithMONVALENTwithPS.eps}
  \caption{\label{PSresponseTONaCl}
    Order parameters of PC (left) and PS (right) headgroup as a function of added NaCl measured from POPC:POPS (5:1) mixture \cite{roux90}.
  }
\end{figure*}
The experimental results show essentially no changes in the order parameters as a function of
added NaCl, while significant changes are observed in simulations. However,
the minimum buffer concentration of NaCl in the experimental was 100mM \cite{roux86}.
Therefore, we cannot exclude the possibility that the NaCl induced changes were already
saturated with 100mM NaCl concentration, which was the case for CaCl$_2$ in Fig. \ref{changesWITHCaClPS}.
\todo{It would be probably better to use datasets for NaCl and KCl from Ref. \citenum{roux90}
to compare monovalent ion binding to PS containing bilayer.}

@@

Fig. \ref{HGorderparametersPCvsPS} shows the PC and PS headgroup order parameters
from POPC:POPS mixtures with various mole fractions from 
simulations and experiments \cite{scherer87,roux90}.
The experimentally observed increase of PC headgroup order parameters with the increasing
amount of negatively charged PS lipid is reproduced in
MacRog simulations with potassium counterions, but not in CHARMM36 simulations
with potassium or sodium conterions. The results can be
understood from the counterion density distributions shown in Fig. \ref{CIdensPSOCmixt}.
Both counterions are more strongly bound to bilayers in
CHARMM36 simulations than potassium in MacRog simulation.
The cation binding probably neutralizes the effect of negatively
charged lipids in CHARMM36 simulations, while the
weaker binding of potassium in MacRog simulations enables the
increase of order parameters with the added amount of PS.

@@

\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPCvsPS.eps}
  \caption{\label{HGorderparametersPCvsPS}
    Headgroup order parameters from PC:PS mixtures from 
    different simulation models and experiments.
    Left panel shows the PC headgroup order parameters (experimental results from Ref. \citenum{scherer87},
    signs are determined as discussed in \cite{botan15,ollila16}).
    Rigth panel shows PS headgroup order parameters (experimental result for pure POPS measured in this
    work at 298K, experimental result for mixture from Ref. \citenum{roux90} at 298K).
    Counterions in expriments are sodium, while potassium is used in simulations.
    Using sodium in simulations do not have a significant effect.
  }
  \todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.} \\
  \todo{We need results also from other than CHARMM36 force field.}
\end{figure*}

\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/CIdensPSOCmixt.eps}
  \caption{  Counterion density distributions from PC:PS mixtures.
\label{CIdensPSOCmixt}
  }
\end{figure}

Also the headgroup order parameters of POPS mixed with varying amounts of POPC
from simulations and experiments \cite{borle85,roux90} are shown in Fig. \ref{HGorderparametersPCvsPS}.
The $\beta$-carbon order parameter of POPS slightly decreases and positive 
order parameter of $\alpha$-carbon slightly increases with increasing
amount of PS lipids. This may indicate increasing order of the headgroup.
It should be, however, noted that the experimental data for pure POPS and
POPC/POPS mixture come from different experimental sets, 13C NMR in this work
and 2H NMR from Ref. \citenum{roux90}, respectively. Therefore the accuracy
of the order parameter change is not as high as typically in the measurements
of order parameter changes, see discussion about qualitative and quantitative 
accuracy in Ref. \citenum{ollila16}. The changes of PS headgroup order parameters
are not reproduced by the tested simulation models. The $\beta$-carbon order parameter
increase with increasing amount of PS in CHARMM36 and MacRog simulations in contrast
to the experimental data. The smaller $\alpha$-carbon order parameter increase in both
simulation models with increasing amount of PS, while it is almost unchanged in experiments.
The larger $\alpha$-carbon order parameter increase in MacRog and decrease in CHARMM36
with increasing amount of PS, both model exhibiting a poor agreement with experiments. 
Signifincant improvement in the MD simulation models are needed to interpret the
PS headgroup structures and mutual interactions with PC lipids.

@@

\begin{figure*}[]
  \centering
  \includegraphics[width=18cm]{../Figs/CHANGESwithCaClPS.eps}
  \caption{\label{changesWITHCaClPS}
    POPC (left) and POPS (right) headgroup order parameters in lipid bilayer with POPC:POPS (5:1) mixture
    as a function CaCl$_2$ concentration from experiments \cite{roux90} and different simulations.
  }
\end{figure*}
\begin{figure}[]
  \centering
  \includegraphics[width=9cm]{../Figs/CAdensPCPSmixture.eps}
  \caption{\label{CAdensPCPSmixture}
    Ca2+ density profiles from simulations.
  }
  \todo{Ca2+ binding in CHARMM is lower than to neutral membrane. The reason should be found out (probably equilibration).}
  \todo{These are now mass densities, numbers would be probably better.}
  \todo{Counterions should be also included.}
  \todo{Not all the data from MacRog is included.}
\end{figure}

@@

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/PSresponseTONaCl.eps}
  \caption{\label{PSresponseTONaCl}
    Order parameters of PS headgroup as a function of added NaCl measured from DMPC:DMPS (3:1) mixture \cite{roux86}.
  }
\end{figure}

>>>>

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/comparisonTablePS.pdf}
  \caption{\label{comparisonTablePS}
  Rough subjective ranking of force fields based on Figure~\ref{HGorderParametersPS}. Here ÒMÓ indicates a magnitude problem, ÒFÓ a forking problem; letter size increases with problem severity. Color scheme: Òwithin experimental errorÓ (dark green), Òalmost within experimental errorÓ (light green), Òclear deviation from experimentsÓ (light red), and Òmajor deviation from experimentsÓ (dark red). The $\Sigma$-column shows the total deviation of the force field, when individual carbons are given weights of 0 (matches experiment), 1, 2, and 4 (major deviation). For full details of the assessment, see Supplementary Information.
  }
  \todo{Issue about possible updates to this plot: https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/4} \\
  \todo{Lipid17 and MacRog results should be added into this plot.}
\end{figure}

The headgroup order parameters of DOPS and POPS bilayers 
from different simulation models are compared with the experimental
data in Fig. \ref{HGorderParametersPS}. In line with the previous
study for PC lipids \cite{botan15}, the glycerol backbone order 
parameters in CHARMM36 roughly agree with the experimental data,
while significant discrepancies for glycerol backbone carbons
are observed in other models.
\todo{Discussion to be finished once all the results are in the plot.} \\
\todo{Also the discussion about POPS/OPPS issue with MacRog model should be added.}

@@

In order to compare the sodium interaction with PS containing membranes between
experiments and simulations, Fig. \ref{PSresponseTONaCl} shows the headgroup
order parameters of POPC and POPS measured from 5:1 mixture as a
function of different monovalent ions added to the buffer.
The experimental experimental results \cite{roux86} are compared with different simulation
models.
\begin{figure*}[]
  \centering
  \includegraphics[width=18.0cm]{../Figs/CHANGESwithMONVALENTwithPS.eps}
  \caption{\label{PSresponseTONaCl}
    Order parameters of PC (left) and PS (right) headgroup as a function of
    added NaCl, KCl and LiCl measured from POPC:POPS (5:1) mixture \cite{roux90}.
  }
  \todo{The upcoming results from Lipid17 have been mentioned in the blog:
    http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1522151331118\#c5876812436739342658} \\
  \todo{Maybe we should get the CHARMM results as well?}
\end{figure*}
In line with the trend observed for bilayers in the absence of negatively
charged lipids \cite{??}, the POPC headgroup order parameter response to lithium ions is
significantly larger than to potassium, indicating stronger binding of lithium
ions to lipid bilayers. The unique nature of lithium is also evident in the
response of PS headgroup order parameters to bounds ions, which decrease
with the addition of lithium or divalent cations but increase with the
addition of sodium or potassium. The headgroup order parameter response
to the added NaCl in the Berger model is similar to the experiments of LiCl,
except for the $\alpha$ carbon of POPS for which the change is overestimated
with the order of magnitude. The results suggests that the binding affinity
of sodium is overestimated by the Berger model, in line with the previous results \cite{catte16},
and that the response of $\alpha$ carbon of POPS to the added sodium
is significantly overestimated. The response of POPC to the added potassium
in MacRog model is only slightly overestimated, in contrast to the response
of POPS where order of magnitude overestimations of reposonse are observed. 
In conclusion, it seems that the response of PS headgroup to bound cations
is not correctly captured in the tested models.
\todo{Finish the discussion when more models are available}

\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/CIdensPSOCmixt.eps}
  \caption{  Counterion density distributions from PC:PS mixtures.
\label{CIdensPSOCmixt}
  }
  \todo{Add Berger results.}
\end{figure}

@@

Fig. \ref{HGorderparametersPCvsPS} shows the PC and PS headgroup order parameters
from POPC:POPS mixtures with various mole fractions from 
simulations and experiments \cite{scherer87,roux90}.
The experimentally observed increase of PC headgroup order parameters with the increasing
amount of negatively charged PS lipid is reproduced in
MacRog simulations with potassium counterions, but not in CHARMM36 simulations
with potassium or sodium conterions. The results can be
understood from the counterion density distributions shown in Fig. \ref{CIdensPSOCmixt}.
Both counterions are more strongly bound to bilayers in
CHARMM36 simulations than potassium in MacRog simulation.
The cation binding probably neutralizes the effect of negatively
charged lipids in CHARMM36 simulations, while the
weaker binding of potassium in MacRog simulations enables the
increase of order parameters with the added amount of PS.
\todo{This discussion will be finalized when we have the
results for monovalent ions also from other simulations.}

@@

\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPCvsPS.eps}
  \caption{\label{HGorderparametersPCvsPS}
    Headgroup order parameters from PC:PS mixtures from 
    different simulation models and experiments.
    Left panel shows the PC headgroup order parameters (experimental results from Ref. \citenum{scherer87},
    signs are determined as discussed in \cite{botan15,ollila16}).
    Rigth panel shows PS headgroup order parameters (experimental result for pure POPS measured in this
    work at 298K, experimental result for mixture from Ref. \citenum{roux90} at 298K).
    Counterions in expriments are sodium, while potassium is used in simulations.
    Using sodium in simulations do not have a significant effect.
  }
  \todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.} \\
  \todo{The upcoming results from Lipid17 have been mentioned in the blog:
    http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1522151331118\#c5876812436739342658} \\
  \todo{Maybe we should show the changes also in this plot?}
\end{figure*}

Also the headgroup order parameters of POPS mixed with varying amounts of POPC
from simulations and experiments \cite{borle85,roux90} are shown in Fig. \ref{HGorderparametersPCvsPS}.
The $\beta$-carbon order parameter of POPS slightly decreases and positive 
order parameter of $\alpha$-carbon slightly increases with increasing
amount of PS lipids. This may indicate increasing order of the headgroup.
\todo{The existing literature about this should be mentioned/discussed.}
It should be, however, noted that the experimental data for pure POPS and
POPC/POPS mixture come from different experimental sets, 13C NMR in this work
and 2H NMR from Ref. \citenum{roux90}, respectively. Therefore the accuracy
of the order parameter change is not as high as typically in the measurements
of order parameter changes, see discussion about qualitative and quantitative 
accuracy in Ref. \citenum{ollila16}. The changes of PS headgroup order parameters
are not reproduced by the tested simulation models. The $\beta$-carbon order parameter
increase with increasing amount of PS in CHARMM36 and MacRog simulations in contrast
to the experimental data. The smaller $\alpha$-carbon order parameter increase in both
simulation models with increasing amount of PS, while it is almost unchanged in experiments.
The larger $\alpha$-carbon order parameter increase in MacRog and decrease in CHARMM36
with increasing amount of PS, both model exhibiting a poor agreement with experiments. 
Signifincant improvement in the MD simulation models are needed to interpret the
PS headgroup structures and mutual interactions with PC lipids. 
\todo{This discussion should be updated.}

@@

\begin{figure*}[]
  \centering
  \includegraphics[width=18cm]{../Figs/CHANGESwithCaClPS.eps}
  \caption{\label{changesWITHCaClPS}
    POPC (left) and POPS (right) headgroup order parameters in lipid bilayer with POPC:POPS (5:1) mixture
    as a function CaCl$_2$ concentration from experiments \cite{roux90} and different simulations.
  }
  \todo{Upcoming simulations with original CHARMM36 have been mentioned in the blog:
  http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1520090718976\#c5569269391707740056} \\
  \todo{Upcoming Lipid17 simulations have been mentioned in the blog
    http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1515177306419\#c994825612316235467}
\end{figure*}
\begin{figure}[]
  \centering
  \includegraphics[width=9cm]{../Figs/CAdensPCPSmixture.eps}
  \caption{\label{CAdensPCPSmixture}
    Ca2+ density profiles from simulations.
  }
  \todo{Ca2+ binding in CHARMM is lower than to neutral membrane. The reason should be found out (probably equilibration).}
  \todo{These are now mass densities, numbers would be probably better.}
  \todo{Counterions should be also included.}
  \todo{Not all the data from MacRog is included.}
\end{figure}

@@

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/PSresponseTONaCl.eps}
  \caption{\label{PSresponseTONaClDMPC}
    Order parameters of PS headgroup as a function of added NaCl measured from DMPC:DMPS (3:1) mixture \cite{roux86}.
  }
\end{figure}
The experimental results show essentially no changes in the order parameters as a function of
added NaCl, while significant changes are observed in simulations. However,
the minimum buffer concentration of NaCl in the experimental was 100mM \cite{roux86}.
Therefore, we cannot exclude the possibility that the NaCl induced changes were already
saturated with 100mM NaCl concentration, which was the case for CaCl$_2$ in Fig. \ref{changesWITHCaClPS}.
